# Getting Started

As this site aims to serve as a practical guide to the technical details of these machine learning concepts, we have provided example code in each section that will allow you to take the publicly available data and models and reproduce the figures you see as you explore each section. Below, we will outline how to set your environment up so that you can code along with the material.

## The GitHub Repository

If you are familiar with git and GitHub, you can clone [the repository](https://github.com/nspies13/practical-machine-learning-in-the-clinical-laboratory) to your local machine.

```bash

## Set Up Our Environment

```{r}
## Load Libraries
library(tidyverse)
library(tidymodels)
library(vetiver)
library(pins)
tidymodels_prefer()

## Configure Environment
set.seed(12345)

```

## Load the Example Data

```{r}

## Download and Import BMP Data Directly from FigShare 
#data <- read_csv("https://figshare.com/ndownloader/files/45355549")
#model_realtime <- read_rds("https://figshare.com/ndownloader/files/45451684") |> pluck("model") |> bundle::unbundle()
#model_retrospective <- read_rds("https://figshare.com/ndownloader/files/45451717") |> pluck("model") |> bundle::unbundle()

## Download Locally, Then Import. ** Change These Paths To Your Own Local Downloads **
#data <- arrow::read_feather("../data/anonymized_bmp_data_all.feather")

## Split Data into Training and Validation Sets
#train <- data |> dplyr::filter(dataset == "WashU Train")
#validation <- data |> dplyr::filter(dataset == "WashU Validation")

## Load the Relevant Models
#model_realtime <- read_rds("../data/normal_saline_XGB_BMP_current_with_deltas.rds") |> pluck("model") |> bundle::unbundle()
#model_retrospective <- read_rds("../data/normal_saline_XGB_BMP_prior_current_and_post.rds") |> pluck("model") |> bundle::unbundle()


```

